---
title: "Exploratory Data Analysis"
author: "Jessica Foster"
date: "2023-03-05"
output: html_document
---

_This document will follow Roger D. Peng's Exploratory Data Analysis Checklist (https://bookdown.org/rdpeng/exdata/exploratory-data-analysis-checklist.html) to explore the charlestonManumissions dataset from the History 8510 repository._

## Exploratory Data Analysis Checklist

Let's load all the libraries we'll use in this worksheet: 
```{r message=FALSE, warning=FALSE}
library(ggplot2) 
library(tidyverse)
library(DigitalMethodsData)
library(forcats)
```

### Step 1: Formulation your question

> What were the most common motives for manumissions in the Charleston District during this period?

### Step 2: Read in your data

```{r}
# Load dataset
data("charlestonManumissions")
```

### Step 3: Check the packaging

```{r}
# How many rows (i.e. observations) does the dataset have?
nrow(charlestonManumissions)
```
> The answer matches the number of observations listed in Environment.

```{r}
# How many columns (i.e. variables) does the dataset have?
ncol(charlestonManumissions)
```
> The answer matches the number of variables listed in Environment.

### Step 4: Run str()

```{r}
# Examine the classes of each of the columns to make sure they are correctly specified (i.e. numbers are numeric and strings are character, etc.). 
str(charlestonManumissions)
```
> Name, Race, Owner, Motive, Notes variables are characters/strings, which makes sense. It would be nice if the Date variable was broken into separate columns for year, month, and date, each being an integer. That would make it easier to perform calculations. Right now Date is a character/string with a combination of numbers and letters. Page variable is also a character even though the first values look like integers. However, page numbers can contain letters.

### Step 5: Look at the top and the bottom of your data

```{r}
# Look at the first 10 rows of data
head(charlestonManumissions, n=10L)
```
> This shows me that the Page variable is a character for a reason, because it is sometimes a range, e.g. 585-86. Looking at the Name column, I can see that each observation is a manumission event, which means there may be more than one person manumitted at the same time. It's unclear whether the names are always comma separated. There are some NA values.

```{r}
# Look at the last 10 rows of data
tail(charlestonManumissions, n=10L)
```
> This shows more NA values under Race. Also I see that some entries under Motive have more than one value listed, separated by a semicolon. I may need to use grepl() to filter for particular Motives.

### Step 6: Check your "n"s

```{r}
# Check the Race variable for spelling variations and/or data entry errors
table(charlestonManumissions$Race)
```
> I see that there are many different spellings for the historical term Mestizo: "Mustizo", "Mustizoe", "Mustee". There are some entries with multiple racial identifiers listed and many NA's. This could be due to the fact that these are manumission events involving multiple enslaved individuals.

```{r}
# Let's check the Motive variable see how it breaks down
table(charlestonManumissions$Motive)
```
> Some of the terms for Motive are unclear. What does it mean to have a "Foreign" motive? "Restricted"? "Sexual"? Also this confirms that more than one motive is listed for some entries. This could be due to the fact that these are manumission events involving multiple enslaved individuals.

```{r}
# To find out what some of these unclear Motives might mean, let's look at the Notes column
manumission.motives <- charlestonManumissions %>%
  select(Name, Owner, Motive, Notes...Quotes) %>%
  filter(grepl('Sexual|Restricted|Foreign', Motive))
```

> Ok so the "Sexual" motive seems to indicate the manumission of the owner's children resulting from sexual unions with enslaved women. For example, Gabriel Powell manumitted a young boy named Jeffrey Powell who was the son of his "mulatto woman named Nancy". This implies Jeffrey is the offspring of the owner's sexual union with an enslaved woman. Powell's motive is also listed as "Moral" because he mentions his wishes for the future well being of the boy.

> For "Foreign," it looks like that involves the manumission of enslaved individuals outside America. An example would be Thomas Lamley of Nassau in the Bahamas manumitting an enslaved woman named Phoebe, or Jerome Laval of Port-au-Prince (Haiti).

> "Restricted" means the manumission has conditions. For example, the enslaved individual will be freed only when they reach a certaiin age. In my opinion, that's more a type of manumission rather than a motive. The same goes for "Outside Purchase", which is not really a motive but a description of how the enslaved individual was acquired.

### Step 7: Validate with at least one external data source

> I would want to look at additional datasets of manumission records from other regions in South Carolina and possibly other states to see if the Motives assigned to this dataset make sense within the larger context of slavery in the South.

### Step 8: Try the easy solution first

> Return to question: What were the most common motives for manumissions in the Charleston District during this period?

```{r}
# Let's go back to the dataset, group according to Motive, and summarize the results
manumission.motives <- charlestonManumissions %>%
  group_by(Motive) %>%
  summarize(count = n())

# We can do a quick visualization of this data with a barplot
ggplot(manumission.motives, aes(x=Motive, y=count)) + 
  geom_bar(stat = "identity") +
  coord_flip() # Let's make it horizontal we can read the labels
```
> Since the bars are not in descending order, it's hard to see which Motives are the most common after "Moral". Time to make a new graph.

```{r}
manumission.motives %>%
  mutate(Motive = fct_reorder(Motive, count)) %>% # Let's reorder the Motive groups from largest to smallest
  ggplot( aes(x=Motive, y=count)) +
    geom_bar(stat="identity") +
    coord_flip() # Let's make it horizontal we can read the labels
```
> This graph shows that the five most common motives for manumission are:
1. Moral
2. Self-Purchase
3. Moral; Sexual
4. Outside Purchase
5. Moral; Foreign

### Step 9: Challenge your solution

> The problem with grouping by Motive is that each manumission event can have multiple Motives listed, so you can't always single out a particular one to say that it's the most common. The above graph shows that "Moral" is the most common motive, but it's often combined with others. The same goes for "Self-Purchase", which is combined with many other motives. Thus, the barplot reflects the most common Motive groupings, not the most common single Motives. That said, humans rarely act from a single motive, so the groupings are more reflective of "real life" and the data is still historically valuable. It will require more work with the dataset to tease out the frequency of the individual Motives. See below.

```{r}
manumission.motives <- manumission.motives %>%
  mutate(Motive = ifelse(Motive == "", NA, Motive)) %>% # Replace empty string with NA
  mutate(Motive = ifelse(is.na(Motive), "Unknown", Motive)) %>% # Replace NA with "Unknown"
  separate_rows(Motive, sep = ";", convert = FALSE) %>% # Put each Motive on separate row
  mutate(Motive = str_trim(Motive, side = "left")) %>% # Remove leading whitespace from each row to prevent false uniques
  group_by(Motive) %>% # Group by Motive
  summarize(count = sum(count)) # Merge duplicate Motives and recalculate totals
```

> Now that we've separated out the individual Motives and tallied them up, we can try creating a barplot again to see how the results differ.

```{r}
manumission.motives %>%
  mutate(Motive = fct_reorder(Motive, count)) %>% # Let's reorder the Motive groups from largest to smallest
  ggplot( aes(x=Motive, y=count)) +
    geom_bar(stat="identity") +
    coord_flip() # Let's make it horizontal we can read the labels
```
> It is now much easier to identify the five most common Motives for manumissions:
1. Moral
2. Self-Purchase
3. Outside Purchase
4. Sexual
5. Foreign

> The first two Motives are unchanged from the previous ranking. However, the third through fifth have changed. Outside Purchase has moved from fourth to third most common (again, it's not really a motive, but for the sake of this exercise, we can treat it as one). Sexual has moved down to fourth, whereas it used to be third when combined with Moral. Foreign is fifth whether or not it is grouped with Moral.

> The Motive groupings are still important since, as I said above, humans rarely act from one motive alone. Thus it's meaningful to see that the third most common group is Moral;Sexual. But now we can see how often each single Motive appears in the entries. I think it's best to use both graphs.

### Step 10: Follow up

1. Do you have the right data?

> I would say that this dataset is NOT enough to make a formal argument in, say, a research paper, about the motives behind manumissions. While the information is valuable, the way the Motive categories were applied is problematic. For one thing, not all the Motives are, well, motives. Some would be better called a type of manumission (e.g. Restricted). Others are more applicable to the history of enslavement for the individual(s) being manumitted (e.g. Outside Purchase) or the context of the manumission (e.g. Foreign). The term "Sexual" is also problematic because it more accurately describes the relationship between the slave owner (usually male) and the enslaved person (usually female). I'm not sure you could say that the slave owner acts out of "sexual" motives when they free the offspring of their sexual union with an enslaved person. It too describes the enslavement.

> That said, a re-creation of the dataset from the original sources using a better system would make it more reliable.

2. Do you need other data?

> Yes. I would want to investigate scope of this dataset. The description says it contains "all of the manumissions filed in Charleston District from 1776-1800 contained in the Miscellaneous Records section of the South Carolina Department of Archives and History." Why Miscellaneous? Is this the only section of the SC Dept. of Archives & History that contains manumissions data for Charleston? Are the records after 1800 really no longer extant?

> To make a stronger argument, besides going back to the sources to re-evaluate the Motives categories, I would want to look at additional datasets of manumission records from other regions in South Carolina and possibly other states. Then I could try to identify the motives in those records to see if the ones in charlestonManumissions make sense. Other helpful sources could be letters written by slave owners describing the reasons for freeing the enslaved people in their household.

3. Do you have the right question?

> I think my question is valid and interesting. However, as I stated above, I don't think asking this question of charlestonManumissions in its current form is sufficient. I would need to look at additional manumission records, as well as letters and other documents written by slave owners concerning manumission. Comparing what the slave owner said in the official manumission record to what they wrote privately in a letter to a family member could be interesting. Besides primary sources, it would be helpful to read secondary literature on manumissions to see what motives scholars have identified so far.